---
title: "Basics of ggplot2"
output:
  learnr::tutorial:
    toc_depth: 3
    theme: united
    highlight: pygments
runtime: shiny_prerendered
description: "Learnr tutorial to start off with ggplot2. Created for a course for PhD students at Medical Faculty, Univ.of Bergen."
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
library(learnr)
library(tidyverse)
library(patchwork)
library(here)

knitr::opts_chunk$set(echo = TRUE)

consultations_gp <- read_delim(
  here("tutorials", "Lesson1", "data", "consultations_cleaned_2024-08-13.txt"),
  delim = "\t",
  show_col_types = FALSE
)
consultations_gp_grouped <- consultations_gp %>%
  group_by(year, age)

# calculate sum of patients per year and group
consultations_gp_summary <- consultations_gp_grouped %>%
  filter(diagnosis != "Alle diagnoser") %>%
  summarise(sum = sum(nConsultations, na.rm = TRUE))

my_plot1 <- ggplot(consultations_gp_summary,
        aes(x = as.factor(year), y = sum)) +
	geom_col(aes(fill = age), position = position_dodge())

consultations_gp_median <- consultations_gp_grouped %>%
  summarise(median = median(nConsultations, na.rm = TRUE),
            stderr = sd(nConsultations, na.rm = TRUE),
            .groups = "drop")

my_plot2 <- ggplot(consultations_gp_median,
                   aes(as.factor(year), median)) +
  geom_col(aes(fill = age), position = position_dodge())

final_plot <- ggplot(consultations_gp_grouped,
                      aes(x = as.factor(year), y = nConsultations)) +
	geom_col(aes(fill = age), position = position_dodge()) +
	facet_grid(cols = vars(gender), rows = vars(diagnosis), scales = "free_y")

my_plot1_nicer <- my_plot1 +
  scale_fill_brewer(type = "div", palette = "PRGn",
                    name = "Age group") +
  xlab("Year") +
  ylab("Total number of consultations") +
  labs(title = "Number of consultations at general practicioners (GPs)",
       subtitle = "groupped by age",
       caption = "data source: aggregated data from SSB")
```


## Setup

Load the necessary libraries

```{r}
library(tidyverse)
library(patchwork)
```

## About `ggplot2`

A plot in `ggplot2` consists of several layers:

1. data
    - strict format: tidy data!
    - one row per datapoint
    - all grouping must be included in the data

2. aesthetics (`aes`) ([original documentation](https://ggplot2.tidyverse.org/reference/index.html#section-aesthetics))
    - how to map the data to the graph?
    - which column is *x*, *y*, ...
    - which column provides grouping (based on the grouping, one can either connect points into lines, split a graph into several facets, or color differently each group)

3. `geom`s ([original documentation](https://ggplot2.tidyverse.org/reference/index.html#section-geoms))
    - how to visualize the data?
    - points (`geom_point`), lines (`geom_line`), bars (`geom_bar`), etc.
    - `geom`s are connected to `stat`s that conduct any necessary pre-processing of data (e.g., `geom_histogram` would first calculate the number of observations in each bin through `stat_bin`)

4. `scale`s ([original documentation](https://ggplot2.tidyverse.org/reference/index.html#section-scales))
    - any type of data representation on the plot
    - coordinates: `scale_x_continuous`, `scale_x_discrete`, `scale_x_date`, etc.
    - colors: `scale_color_manual`, `scale_colour_brewer`, `scale_fill_continuous`, etc.

5. `theme` ([original documentation](https://ggplot2.tidyverse.org/reference/index.html#section-themes))
    - visual aspects
    - size and types of fonts
    - positioning of axes, legends, etc.

Each layer is added to the previous one with a `+` sign. The result can be saved to an object and added upon later.

### Example

Let's read in some data.

```{r, eval=FALSE}
consultations_gp <- read_delim(
  here("data", "consultations_cleaned_2024-08-13.txt"),
  delim = "\t"
)
consultations_gp
```

This is publicly available data from SSB (Norwegian Statistics Bureau) on number of consultations at general practitioners (GP) between 2012 and 2021.

Let's plot some numbers: total number of patients per year and per group ("diagnosis").

---

#### Grouping the data

```{r}
consultations_gp_grouped <- consultations_gp %>%
  group_by(year, age)

# calculate sum of patients per year and group
consultations_gp_summary <- consultations_gp_grouped %>%
  filter(diagnosis != "Alle diagnoser") %>%
  summarise(sum = sum(nConsultations))
consultations_gp_summary
```

We see that the data has now only `NA` values - this is because, by default,
the `sum` function does not discard `NA` values. Thus, we need to specify this:

```{r}
consultations_gp_summary <- consultations_gp_grouped %>%
  filter(diagnosis != "Alle diagnoser") %>%
  summarise(sum = sum(nConsultations, na.rm = TRUE)) # don't count NAs
consultations_gp_summary
```


---

#### Plotting    
Here, we're using simplest type of plot: points.

```{r}
ggplot(consultations_gp_summary,  # define data
       aes(x = as.factor(year), y = sum)) +  # define which columns are x and y
  geom_point() # define way to visualize the data
```

> NOTE: we used `as.factor` when defining column `year` as `x` because we want
the x-axis to be categorical and sice in the `year` columns are only numbers,
by default it would be viewed as a continuous variable.

This does not look great - we have several points per year since there are many
categories ("age groups"), but we don't know which point is which.

---

#### Add coloring

Here is the magic of ggplot - we want to add coloring of the points *by one of
the variables in the dataset*. Therefore, we need to specify aesthetics
for the `geom_point`.

```{r}
plot_sum_consult <- ggplot(consultations_gp_summary,
       aes(x = as.factor(year), y = sum)) +
  geom_point(aes(col = age))
plot_sum_consult
```

...and we get a legend for free!

> NOTE: we can save the plots as objects and then add on to those if we want
to modify the plot!

### Exercise 1

Plot the same but now using bar chart instead of points (`geom_col`)

```{r geom-col, exercise=TRUE}

```

```{r geom-col-solution}
ggplot(consultations_gp_summary,
       aes(x = as.factor(year), y = sum)) +
  geom_col(aes(col = age))
```

**What happened to the color?**

To _fill_ the bars with colors, you need to use the `fill` argument of the
aesthetics, not the `col` (color) argument. Try this below!

```{r geom-col-fill, exercise = TRUE}

```

```{r geom-col-fill-solution}
ggplot(consultations_gp_summary,
       aes(x = as.factor(year), y = sum)) +
  geom_col(aes(fill = age))
```

### Exercise 2

Now plot the same using lines:

```{r geom-lines, exercise=TRUE, exercise.lines = 5}
plot_sum_consult_lines <- ggplot(consultations_gp_summary,
       aes(x = as.factor(year), y = sum)) +
  ___(aes(group = age))
plot_sum_consult_lines
```

**What does the `group` argument of the `aes` function do?** *(check by 
experimenting in the box above or copying from `Hint` code)*

```{r geom-lines-hint}
# `group` argument helps draw lines connecting the same categories of points
# run the code below (without the `group` argument) to see how strange the
# result would be
ggplot(consultations_gp_summary,
       aes(x = as.factor(year), y = sum)) +
  geom_line(aes(col = age))
```

```{r geom-lines-solution}
ggplot(consultations_gp_summary,
       aes(x = as.factor(year), y = sum)) +
  geom_line(aes(group = age))
```




## More elements on one plot

### Faceting

We've plotted the number of admissions per year in total and per group.
Sometimes these numbers vary depending on other variables, here age (`age`) or
sex (`gender`). With ggplot2, it's easy to split the plot in different
sub-plots, called _facets_.

```{r, fig.cap="**Figure above:** No splitting"}
# plotting totals, grouped, non-stacked
ggplot(consultations_gp_summary,
        aes(x = as.factor(year), y = sum)) +
	geom_col(aes(fill = age), position = position_dodge())
```


```{r, fig.cap="**Figure above:** Splitting once", out.width = "100%", fig.height = 6, fig.width = 8}
# splitting this - checking difference with regard to gender
ggplot(consultations_gp_grouped,
        aes(x = as.factor(year), y = nConsultations)) +
	geom_col(aes(fill = age), position = position_dodge()) +
	facet_wrap(facets = vars(gender))
```


```{r, fig.cap="**Figure above:** Splitting twice", out.width = "100%", fig.height = 12, fig.width = 8}
# splitting again - with regard to age
final_plot <- ggplot(consultations_gp_grouped,
                      aes(x = as.factor(year), y = nConsultations)) +
	geom_col(aes(fill = age), position = position_dodge()) +
	facet_grid(cols = vars(gender), rows = vars(diagnosis))
final_plot
```

Suddenly, we can't see some of the bars because the heights are so different.
When that happens when we split with `facets`, you can specify which axis can
change independently from group to group by setting the argument `scales`.

```{r, fig.cap="**Figure above:** Splitting twice, free y scale", out.width = "100%", fig.height = 12, fig.width = 8}
final_plot <- ggplot(consultations_gp_grouped,
                      aes(x = as.factor(year), y = nConsultations)) +
	geom_col(aes(fill = age), position = position_dodge()) +
	facet_grid(
	  cols = vars(gender),
	  rows = vars(diagnosis),
	  scales = "free_y"
  )
final_plot
```


### Exercise 1

Plot the data using `age` and `diagnosis` as facets.

```{r plot_facets, exercise=TRUE, exercise.eval=TRUE, exercise.lines = 5, out.width = "100%", fig.height = 12, fig.width = 8}


```

```{r plot_facets-solution, out.width = "100%", fig.height = 12, fig.width = 8}
ggplot(consultations_gp_grouped,
                      aes(x = as.factor(year), y = nConsultations)) +
	geom_col(aes(fill = gender), position = position_dodge()) +
	facet_grid(
	  rows = vars(diagnosis),
	  cols = vars(age),
	  scales = "free_y"
	 )
```

---

### Several geoms

Of course, we can plot several various types of graphs on one plot.

> NOTE: with ggplot, you can add more things also _after_ the plot was created
by saving it to an object and adding onto the object.

```{r}
my_plot1 <- ggplot(consultations_gp_summary,
        aes(x = as.factor(year), y = sum)) +
	geom_col(aes(fill = age), position = position_dodge())

my_plot1 +
  geom_line(aes(group = age))
```

Something is not right - the lines are drawn between the points that are in the
middle of each year-category. This is because we used `position_dodge()`, which
artificially splits the bars around the middle of the category.

So we need to do the same with lines.

```{r}
my_plot1 +
  geom_line(aes(group = age, col = age),
            position = position_dodge(width = 0.9))
```

This is not so useful to plot the same values with two different geoms, but
we can plot errorbars! (`geom_errorbar`)

For that, we need to first construct a dataset that contains e.g., median and
standard error of the inpatients per year.

> *Remember to set the functions to ignore the NAs!*

```{r}
consultations_gp_median <- consultations_gp_grouped %>%
  filter(diagnosis != "Alle diagnoser") %>%
  summarise(median = median(nConsultations, na.rm = TRUE),
            stderr = sd(nConsultations, na.rm = TRUE))
consultations_gp_median
```

<br>

Let's plot the median as bars:

```{r}
my_plot2 <- ggplot(consultations_gp_median,
                   aes(as.factor(year), median)) +
  geom_col(aes(fill = age), position = position_dodge())
my_plot2
```

----

### Exercise 2

Now, your turn - let's add the errorbars!

> HINT: `geom_errorbar` uses `ymin` and `ymax` arguments of `aes`

> HINT (2): use also the `group` parameter of `aes` and `position_dodge`, as
for the `geom_line` example above

```{r add_errorbars, exercise=TRUE, exercise.eval=TRUE, exercise.lines = 5}

```

```{r add_errorbars-solution}
my_plot2 +
  geom_errorbar(
    aes(
      ymin = (median - stderr),
      ymax = (median + stderr),
      group = age
     ),
    position = position_dodge(width = 0.9)
 )
```

---

### Combining different plots

Sometimes, we want to plot several various plots and then combine them into
one figure. This can't be done with `ggplot2` itself, but there are many
additional packages that extend `ggplot2`. One of them is very useful for
combining plots - [`patchwork`](https://patchwork.data-imaginist.com/index.html).

Until now, we have several plots: `final_plot`, `my_plot1`, `my_plot2`. Let's
try putting those together:

```{r, out.width = "100%", fig.height = 6, fig.width = 8}
# next to each other
plot_sum_consult + my_plot1
```


```{r, out.width = "100%", fig.height = 14, fig.width = 8}
# on top of each other
final_plot / my_plot1
```

As easy as that!

Legends are the same, so let's gather those.

```{r, out.width = "100%", fig.height = 14, fig.width = 8}
final_plot / my_plot1 + 
  plot_layout(guides = 'collect')
```


### Exercise 3

Play with patchworking a bit, if you like!

- make ggplot objects
- combine 4 plots on the same page, e.g.,:

```{r, eval = FALSE}
(my_plot1 + my_plot2) / (my_plot3 + my_plot4)
```

```{r patchwork-fun, exercise = TRUE}

```


## Prettify

Until now, we were trying out:

- plotting with various `geom`s
- using variables to guide coloring
- dodging the points within the same category
- faceting
- combining various plots into one

However, we should not be satisfied with the default visualizations! Color
palettes, axis naming, fonts, adding titles, etc. - all this can be easily
changed in `ggplot2`.

### Color palette

The colors are turned on and off within the `aes` function. However, the
_palette_ can be changed independently. There is a family of functions that
control this, starting with `scale_`.

> One issue that you should keep in mind when choosing the palette is that
they can change our perception of displayed data. Moreover, one needs to 
stick to palettes that increase the accessibility (e.g., for the color-blind,
or those with sight problems). There is a very good publication that I
recommend: https://www.nature.com/articles/s41467-020-19160-7

There are many different palettes available on the net - we will focus on
those that are available within `ggplot2`.

**Discrete colors**

In the examples above, we're using coloring to show various groups of patients.
Thus, we need a palette that would easily differentiate between the groups.
`ggplot2` uses [ColorBrewer](https://colorbrewer2.org) palettes - on the 
webpage, one can choose and check visually various color sets. Now, we will
use the "PRGn" (purple-green) palette, but you're welcomed to try any other.

```{r, fig.cap="Ah, much nicer!"}
my_plot1 +
  scale_fill_brewer(type = "div", palette = "PRGn")
```

### Exercise 1

Change the colors below, using two different palettes: one for the bars,
and one for the lines. Choose the set of colors from ColorBrewer webpage.

> Fill in the `___` fields!

```{r adjust_palettes, exercise=TRUE, exercise.lines = 7}
my_plot1 +
  geom_line(aes(group = age, col = age),
            position = position_dodge(width = 0.9)) +
  scale_____brewer(type = "div", palette = ___) +
  scale_____brewer(type = "div", palette = ___)
```

```{r adjust_palettes-solution}
my_plot1 +
  geom_line(aes(group = age, col = age),
            position = position_dodge(width = 0.9)) +
  scale_fill_brewer(type = "div", palette = "BrBG") +
  scale_color_brewer(type = "div", palette = "PuOr")
```

### Axes

Let's add some explanations to the plot: axes titles, main titles, legend title.

```{r}
my_plot1 +
  scale_fill_brewer(type = "div", palette = "PRGn") +
  xlab("Year") + # changing the x label
  ylab("Total number of consultations") +
  labs(title = "Number of consultations at general practicioners (GPs)",
       subtitle = "groupped by age",
       caption = "data source: aggregated data from SSB")
```

The legend text is modified within the `scale` function.

```{r}
my_plot1_nicer <- my_plot1 +
  scale_fill_brewer(type = "div", palette = "PRGn",
                    name = "Age group") + # name of the legend
  xlab("Year") +
  ylab("Total number of  consultations") +
  labs(title = "Number of consultations at general practicioners (GPs)",
       subtitle = "groupped by age",
       caption = "data source: aggregated data from SSB")
my_plot1_nicer
```

### Theme

All the visual aspects can be changed within the `theme` command.

For our convenience, there are some pre-defined themes that combine several
aspects, such as line thickness, background color, font families, etc.

### Excercise 2

Check what happens when you add `theme_minimal()` to our `my_plot1_nicer`.

```{r check_theme_min, exercise=TRUE, exercise.lines = 4}

```

```{r check_theme_min-solution}
my_plot1_nicer + theme_minimal()
```

---

To control these visual aspects better, you can check [the documentation](https://ggplot2.tidyverse.org/reference/theme.html). Below, there
are some popular useful examples.

**1. Change the fonts**

```{r}
my_plot1_nicer +
  theme(
    plot.title = element_text(
      size = 20, # text size in pts
      face = "bold"
   ),
    axis.title = element_text(
      size = 16
   ),
    axis.text.x = element_text(
      angle = 90
   )
 )
```

**2. Change the position of the legend**

```{r}
my_plot1_nicer +
  theme(
    legend.position = "bottom"
 )
```

**3. Change the background color**

```{r}
my_plot1_nicer +
  theme(
      panel.background = element_rect(fill = "salmon")
 )
```


### Exercise 3

Now it's your turn - play around!

```{r theme_fun, exercise=TRUE, exercise.lines = 5}

```


